---
title: "Introduction to Formula Response Function Recur()"
author: Wenjie Wang
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
# bibliography: ../inst/bib/reda.bib
vignette: >
  %\VignetteIndexEntry{Introduction to Formula Response Function Recur()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, echo = 3:4}
knitr::opts_chunk$set(comment = "")
head <- utils::head
library(reda)
packageVersion("reda")
```

# Overview

The `Recur()` function provides a flexible and widely applicable formula
response interface for modeling recurrent event data with considerate data
checking procedures.  It combined the flexible interface of `reSurv()`
(deprecated in **reReg** version 1.1.7) and the effective checking procedures
embedded in the `Survr()` (deprecated in **reda** version 0.5.0).


# Function Interface

The function interface of `Recur()` is given below.

```R
Recur(time, id, event, terminal, origin, check = c("hard", "soft", "none"), ...)
```

A high-level introduction to each argument is as follows:

- `time`: event and censoring times
- `id`: subject's id
- `event`: recurrent event indicator, cost, or type
- `terminal`: event indicator of terminal events
- `origin`: time origin of subjects
- `check`: how to run the data checking procedure
  - `"hard"`: throw errors if the `check_Recur()` finds any issue on the data
    structure
  - `"soft"`: throw warnings instead
  - `"none"`: not to run the checking procedure

More details of arguments are provided in the function documentation by
`?Recur`.


# The `Recur` Object

The function `Recur()` returns an S4-class `Recur` object representing model
response for recurrent event data.  The `Recur` class object mainly contains a
numerical matrix object (in the `.Data` slot) that serves as a model response
matrix.  The other slots are

- `ID`: a factor storing the original subject's ID, which originally can be a
  character vector, a numeric vector, or a factor).  It is needed to pinpoint
  data issues for particular subjects with their original ID's.
- `ord`: indices that sort the response matrix (by rows) increasingly by `id`,
  `time2`, and `- event`.  Sorting is often done in the model-fitting steps,
  where the indices stored in this slot can be used directly.
- `rev_ord`: indices that revert the increasingly sorted response matrix by
  `ord` to its original ordering.  This slot is provided to easily revert the
  sorting.
- `first_idx`: indices that indicates the first record of each subject in the
  sorted matrix.  It helps in the data checking produce and may be helpful in
  model-fitting step, such as getting the origin time.
- `last_idx`: indices that indicates the last record of each subject in the
  sorted matrix.  Similar to `first_idx`, it helps in the data checking produce
  and may be helpful in the model-fitting step, such as locating the terminal
  events.
- `check`: a character string that records the specified `check` argument.  It
  just records the option that users specified on data checking.


# Usage

Among all the arguments, only the argument `time` does not have default values
and thus has to be specified by users.


## When only `time` is given

- The function assumes that each time point is specified for each subject.
- The `id` takes its default value: `seq_along(time)`.
- The `event` takes its default values: `0` (censoring) at the last record of
  each subject, and `1` (event) before censoring.
- Both `terminal` and `origin` take zero for all subjects by default.

```{r ex-1}
ex1 <- Recur(3:5)
head(ex1)
```


## When `time` and `id` are given

- The `event` takes its default values: `0` (censoring) at the last record of
  each subject, and `1` (event) before censoring.
- Both `terminal` and `origin` take zero for all subjects by default.

```{r ex-2}
ex2 <- Recur(6:1, id = rep(1:2, 3))
head(ex2)
## sort by id, time2, and - event
head(ex2[ex2@ord, ])
```

- The slot `ord` stores the indices that sort the response matrix by `id`,
  `time2`, and `- event`.


## Helper `%to%` for recurrent episodes

The function `Recur()` allows users to input recurrent episodes by `time1` and
`time2`, which can be specified with help of `%to%` (or its alias `%2%`) in
`Recur()`.  For example,

```{r ex-3}
left <- c(1, 5, 7)
right <- c(3, 7, 9)
ex3 <- Recur(left %to% right, id = c("A1", "A1", "A2"))
head(ex3)
```

Internally, the function `%to%` returns a list with element named `"time1"` and
`"time2"`. Therefore, it is equivalent to specify such a list.

```{r ex-4}
ex4 <- Recur(list(time1 = left, time2 = right), id = c("A1", "A1", "A2"))
stopifnot(all.equal(ex3, ex4))
```


## About `origin`  and `terminal`

- Both `origin` and `terminal` take a numeric vector.
- The length of specified vector can be one, equal to the number of subjects, or
  the number of `time`.  Some simple examples are given below.

```{r ex-5}
ex5 <- Recur(3:5, origin = 1, terminal = 1)
head(ex5)
```

```{r ex-6}
ex6 <- Recur(3:5, id = c("A1", "A1", "A2"), origin = 1:2, terminal = c(0, 1))
head(ex6)
```

```{r ex-7}
ex7 <- Recur(3:5, id = c("A1", "A1", "A2"),
             origin = c(1, 1, 2), terminal = c(0, 0, 1))
stopifnot(all.equal(ex6, ex7))
```

- An error message will be thrown out if the length is inappropriate.

```{r origin-terminal-err}
try(Recur(1:10, origin = c(1, 2)))
try(Recur(1:10, terminal = c(1, 2)))
```


## Data Checking Rules

The `Recur()` (internally calls `check_Recur()` and) checks whether the
specified data fits into the recurrent event data framework by several rules if
`check = "hard"` or `check = "soft"`.  The existing rules and the corresponding
examples are given below.

1. Every subject must have one censoring not before any event time.

```{r rule1}
try(Recur(1:5, id = c(rep("A1", 3), "A2", "A3"), event = c(0, 0, 1, 0, 0)))
```

2. Every subject must have one terminal event time.

```{r rule2}
try(Recur(1:3, id = rep("A1", 3), terminal = c(0, 1, 1)))
```

3. Event or censoring times cannot be missing.

```{r rule4}
try(Recur(c(1:2, NA), id = rep("A1", 3)))
```

4. Event times cannot be earlier than the origin time.

```{r rule5}
try(Recur(3:5, id = rep("A1", 3), origin = 10))
try(Recur(3:5 %to% 1:3, id = rep("A1", 3)))
```

5. The recurrent episode cannot be overlapped.

```{r rule6}
try(Recur(c(0, 3, 5) %to% c(1, 6, 10), id = rep("A1", 3)))
```

6. However, recurrent episode without events is allowed for possible
   time-varying covariates and risk-free gaps.

```{r rule3}
Recur(c(0, 2, 6) %to% c(1, 3, 8), id = rep("A1", 3), event = c(0, 1, 0))
```


# The `Show()` Method

A `show()` method is added for the `Recur` object in a similar fashion to the
output of the function `survival:::print.Surv()`, which internally converts the
input `Recur` object to character strings representing the recurrent episodes by
a dedicated `as.character()` method.  For each recurrent episode,

- Censoring not due to terminal is indicated by a trailing `+` sign;
- Censoring due to terminal is indicated by a trailing `*` sign;
- Otherwise, an event happens at the end of the recurrent episode.

For a concise printing, the `show()` method takes the
`getOption("reda.Recur.maxPrint")` to limit the maximum number of recurrent
episodes to be printed for each process.  By default,
`options(reda.Recur.maxPrint = 3)` is set.


## The Valve Seats Example

We may illustrate the results of the `show()` method by the example valve seats
data, where terminal events are artificially added.

```{r show1}
set.seed(123)
term_events <- rbinom(length(unique(valveSeats$ID)), 1, 0.5)
with(valveSeats, Recur(Days, ID, No., term_events))
```

## On Missing times

The updated `show()` method preserves `NA`'s when `check = "none"`.  However,
`NA`'s will always appear because times are sorted internally.

```{r show-na}
Recur(c(NA, 3:6, NA), id = rep(1:2, 3), check = "none")
```

## Number of digits

The `show()` method takes the value of `options("digits") - 3` to determine the
largest number of digits for printing.

```{r show-dig1}
op <- options()
getOption("digits")
Recur(pi, 1)
```

```{r show-dig2}
options(digits = 10)
Recur(pi, 1)
options(op) # reset (all) initial options
```



