---
title: "Simulate Survival and Recurrent Event Data with reda"
author: Wenjie Wang
date: "`r Sys.Date()`"
bibliography:
- ../inst/bib/reda.bib
vignette: >
  %\VignetteIndexEntry{Simulate Survival and Recurrent Event Data with reda}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
output: rmarkdown::html_vignette
---


In this vignette, we briefly introduce how to simulate survival and recurrent
event data from a stochastic process point of view with **reda** package.  The
core function is named `simEve` for simulating survival and recurrent event
times for one stochastic process. Another function named `simEveData` is simply
a wrapper that calls `simEve` internally and collects event times and covariates
of a given number of processes into a data frame format.  Examples for
generating random samples from common survival and recurrent models are
provided. The details of function syntax and the objects produced are available
in the package manual and thus not covered in this vignette.


## Hazard Rate Function

We first introduce the general form of hazard rate function that was considered
for the function `simEve`.  Let $\rho(t)$ denote the baseline hazard rate
function of time $t$, $\mathbf{x}_i(t)$ denote the covariate vector,
$\boldsymbol{\beta}(t)$ denote the covariate coefficients for a stochastic
process $i$. Note that both covariates and coefficients can be time-varying.
The @andersen1982aos model incorporates covariates and their coefficients into
the stochastic process by specifying the intensity as a function of time $t$ and
the covariates,
$$
\lambda(t \mid \mathbf{x}_i, r_i) = r_i\rho(t)
\exp\left\{\mathbf{x}^{\top}_i(t) ~\boldsymbol{\beta}(t)\right\},
$$
where


## Homogenous Poisson Process {#hpp}

```{r setup}
library(reda)
```

```{r hpp}
simEve(z = c(0.2, 0.5), zCoef = c(1, - 0.5), rho = 0.5, endTime = 10)
```

```{r hpp.data}
simEveData(z = c(0.2, 0.5), zCoef = c(1, - 0.5), rho = 0.5, endTime = 10)
```




## Reference
